Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-27 _tex/paper2.tex
Built site for gh-pages

Nevertheless, despite the growing number of CJ studies, unsystematic and
fragmented research approaches have left several critical issues
unresolved. This study addresses two such issues that can compromise the
reliability and validity of CJ's trait estimates. The first concerns the
overreliance on Thurstone's Case V assumptions in the statistical
analysis of CJ data. The second focuses on the apparent disconnect
between CJ's approach to trait measurement and hypothesis testing.

The study is organized into six main sections. It begins with an
overview of Thurstone's theory in Section~\ref{sec-thurstone_theory},
followed by a detailed examination of these issues in
Section~\ref{sec-theory-issues}. In Section~\ref{sec-theoretical}, the
study develops a theoretical model for CJ that simultaneously addresses
these issues. The model integrates Thurstone's core principles with key
assessment design features relevant to CJ experiments, such as the
selection of judges, stimuli, and comparisons.
Section~\ref{sec-statistical} translates these theoretical and practical
elements into a statistical model that supports the analysis of pairwise
comparison data. Finally, Section~\ref{sec-discussion} discusses the
concepts developed in the study, their implications, and avenues for
future exploration, while Section~\ref{sec-conclusion} summarizes the
findings.

@@

\section{The two critical issues in CJ
literature}\label{sec-theory-issues}

This section explores the two critical issues in CJ's theory that have
the potential to compromise the reliability and validity of CJ's trait
estimates. Section~\ref{sec-theory-issue1} inspects the overreliance on
Thurstone's Case V assumptions in the statistical analysis of CJ data.
Section~\ref{sec-theory-issue2} focuses on the apparent disconnect
between CJ's approach to trait measurement and hypothesis testing.

@@

Thurstone noted from the outset that the general form of the theory, as
outlined in Section~\ref{sec-thurstone_theory}, gave rise to a problem
of trait scaling. The model required estimating more ``unknown''
parameters than the available pairwise comparisons
\citep[pp.~267]{Thurstone_1927b}. To address this issue and facilitate
the practical implementation of the theory, he developed five cases
derived from this general form, each case progressively incorporated
additional simplifying assumptions into the model.

@@

In aggregate, researchers' inadequate handling of violations to the
assumptions of equal dispersion and zero correlation between stimuli,
coupled with the apparent disconnect between CJ's approach to trait
measurement and hypothesis testing, can potentially compromise the
reliability of the trait estimates and, by extension, their validity
\citep[pp.~2]{Perron_et_al_2015}. Consequently, adopting a more
systematic and integrated approach to handling these assumptions and
examining the factors influencing CJ experiments could offer several
statistical and measurement benefits, including the ability to address
the aforementioned issues.

@@

\subsection{The population model}\label{sec-theory-theoretical_P}

@@

Ideally, researchers then proceed to identify the estimands.
\emph{Identification} determines whether an estimator can accurately
compute the estimand based solely on its assumptions, regardless of
random variability \citep[pp.~4]{Schuessler_et_al_2023}. Identification
is crucial because if an estimator cannot reliably compute an estimand
using ``infinite'' and error-free data, it is impossible to gain
meaningful insights about that estimand using finite data
\citep[pp.~5]{Schuessler_et_al_2023}. An \emph{estimator} refers to a
method or function that transforms data into an estimate
\citep{Neal_2020}. \emph{Estimates} are numerical values that
approximate the estimand derived through the process of
\emph{estimation}, which integrates data with an estimator
\citep{Everitt_et_al_2010}. The Identification-Estimation flowchart
\citep{McElreath_2020, Neal_2020}, shown in Figure~\ref{fig-IEflow},
visually represents the transition from estimands to estimates.

@@

Consequently, although many estimators can approximate an estimand,
researchers prioritize those with desirable identification properties to
ensure accurate and reliable estimates. For instance, the Z-test is a
widely used estimator for comparing group proportions, providing
accurate estimates when its underlying assumptions are satisfied
\citep{Kanji_2006}. Moreover, researchers can interpret the Z-test
estimates as causal, provided they collect the data through a randomized
experiment.

@@

Even though counterfactuals are unobservable, researchers can still
identify the ACE from associational estimates by leveraging the
structural approach. The approach identifies the ACE by statistically
conditioning data on a \emph{sufficient adjustment set} of variables
\(X\) \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. This
\emph{sufficient} set (potentially empty) must block all non-causal
paths between \(T\) to \(Y\) without opening new ones. When such a set
exists, then \(T\) and \(Y\) are \emph{d-separated} by \(X\)
(\(T \:\bot\:Y \:|\:X\)) \citep{Pearl_2009}, and \(X\) satisfies the
\emph{backdoor criterion} \citep[pp 37]{Neal_2020}. Here,
\emph{conditioning} describes the process of restricting the focus to
the subset of the population defined by the conditioning variable
\citep[pp.~32]{Neal_2020} (see Equation~\ref{eq-group_diff}).

@@

For instance, If researchers cannot conduct the randomized experiments
described in Section~\ref{sec-appendix-A} and must instead rely on
observational data, they can still identify the ACE as long as an
observed variable \(X\), such as the socio-economic status of the
school, satisfies the backdoor criterion. Under these circumstances,
researchers first identify the \emph{conditional average causal effect}
(CACE, Equation~\ref{eq-CACE})

@@

SCMs and DAGs support causal inference through two key advantages.
First, regardless of complexity, they can represent various causal
structures using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
decompose complex structures into manageable components, facilitating
their analysis \citep{McElreath_2020}. Second, they depict causal
relationships in a non-parametric and fully interactive way. This
flexibility enables feasible ACE identification strategies without
defining the variables' data types, the functional form between them, or
their parameters \citep[pp.~35]{Pearl_et_al_2016}.

Section~\ref{sec-appendix-B21} and Section~\ref{sec-appendix-B22}
elaborate on the first advantage, while Section~\ref{sec-appendix-B22}
and Section~\ref{sec-appendix-B23} do so for the second.

@@

The structural approach to causal inference translate the probabilistic
insights from Section~\ref{sec-appendix-B23} into actionable strategies
seeking to identify the ACE from associational quantities. It achieves
this by relying on the \emph{modularity assumption}, which posits that
intervening on a node alters only the causal mechanism of that node,
leaving others unchanged \citep[pp.~34]{Neal_2020}.

@@

With these new insights, researchers revisiting the scenario in
Section~\ref{sec-appendix-B22} can infer that the socio-economic status
of the school, \(X\), satisfies the backdoor criterion, assuming the
causal structure depicted by the SCM \ref{fig-scm_example2} and DAG
\ref{fig-mdag_example2} is correct. This means that \(X\) serves as a
sufficient adjustment set, as it effectively blocks all confounding
non-causal paths introduced by the fork structure. Nevertheless, since
\(Y\) remains dependent on \(T\) even after conditioning
\((Y \:\not\bot\:T \:|\:X)\), this dependency can only be attributed to
the direct causal effect \(T \rightarrow Y\).

Researchers can then apply the \emph{backdoor adjustment} to identify
the ACE of \(T\) on \(Y\). They achieve this by first identifying the
CACE of \(T\) on \(Y\) by conditioning on \(X\), and then marginalizing
this effect over \(X\) to obtain the ACE. This process is expressed in
Equation~\ref{eq-mACE2} (see Section~\ref{sec-appendix-B}). Notably, for
the purpose of identification, the conditioned DAG
\ref{fig-mdag_example2} is equivalent to the manipulated DAG
\ref{fig-mdag_example3}, because \(X\) satisfies the backdoor criterion.

@@

@@

Nevertheless, researchers can manage the complexity of the likelihood by
assuming that variables exhibit specific local (in)dependencies. These
assumptions enhance model tractability and facilitate the estimation
process. As described in Section~\ref{sec-appendix-B22}, researchers can
use SCMs and DAGs to formalize these local (in)dependencies, which leads
to the BNF of the likelihood (Equation~\ref{eq-like_BNF}). From this
point, any probabilistic programming language can model this simpler
structure and generate the parameter's posterior distribution using
Equation~\ref{eq-bayes_rule}.

>>>>

Nevertheless, despite the increasing number of studies on CJ, research
in this domain remains unsystematic and fragmented, leaving several
critical issues unresolved. This study addresses two prominent issues in
CJ literature that can undermine the reliability and validity of CJ's
trait estimates. The first problem concerns the overreliance on
Thurstone's Case V assumptions in the statistical analysis of CJ data.
The second involves the apparent disconnect between CJ's approach to
trait measurement and hypothesis testing.

The study is organized into six main sections. An overview of
Thurstone's theory is presented in Section~\ref{sec-thurstone_theory},
followed by a detailed examination of the identified issues in
Section~\ref{sec-theory-issues}. In Section~\ref{sec-theoretical}, the
study proposes a theoretical model for CJ that addresses these
challenges. The model incorporates Thurstone's core principles alongside
key assessment design features relevant to CJ experiments, such as
judges, stimuli, and comparison selections. In
Section~\ref{sec-statistical}, these practical and theoretical
considerations are translated into a statistical model for the analysis
of pairwise comparison data. Finally, Section~\ref{sec-discussion}
explores the implications of the study's findings and the avenues for
future exploration, and Section~\ref{sec-conclusion} provides a summary
of the key insights.

@@

\section{Two prominent issues in CJ literature}\label{sec-theory-issues}

This section discusses the two prominent issues in CJ literature that
can undermine the reliability and validity of CJ's trait estimates.
Section~\ref{sec-theory-issue1} examines the overreliance on Thurstone's
Case V assumptions in the statistical analysis of CJ data.
Section~\ref{sec-theory-issue2} focuses on the apparent disconnect
between CJ's approach to trait measurement and hypothesis testing.

@@

Thurstone noted from the outset that the general form of the theory, as
outlined in Section~\ref{sec-thurstone_theory}, gave rise to a trait
scaling problem. Specifically, the model required estimating more
``unknown'' parameters than the available pairwise comparisons
\citep[pp.~267]{Thurstone_1927b}. To address this issue and facilitate
the practical implementation of the theory, he developed five cases
derived from this general form, each progressively incorporating
additional simplifying assumptions into the model.

@@

In aggregate, researchers' overreliance on Thurstone's Case V
assumptions in the statistical analysis of CJ data can compromise the
reliability of CJ's trait estimates. This overreliance, coupled with the
apparent disconnect between CJ's approach to trait measurement and
hypothesis testing, may also undermine the validity of these estimates
\citep[pp.~2]{Perron_et_al_2015}. Consequently, addressing these issues
and examining the factors influencing CJ experiments by adopting a more
systematic and integrated approach could offer several statistical and
measurement benefits.

@@

\subsection{The conceptual population
model}\label{sec-theory-theoretical_P}

@@

Researchers then proceed to identify the estimands.
\emph{Identification} determines whether an estimator can accurately
compute the estimand based solely on its assumptions, regardless of
random variability \citep[pp.~4]{Schuessler_et_al_2023}. An
\emph{estimator} refers to a method or function that transforms data
into an estimate \citep{Neal_2020}. \emph{Estimates} are numerical
values that approximate the estimand derived through the process of
\emph{estimation}, which integrates data with an estimator
\citep{Everitt_et_al_2010}. The Identification-Estimation flowchart
\citep{McElreath_2020, Neal_2020}, shown in Figure~\ref{fig-IEflow},
visually represents the transition from estimands to estimates.

@@

Identification is essential because it is a necessary condition for
\emph{consistency}. An estimator is \emph{consistent} if it converges to
the ``true'' value of an estimand as the sample size approaches infinity
\citep{Everitt_et_al_2010}. Without identification, an estimator cannot
achieve consistency, rendering it incapable of reliably computing the
estimand, even with ``infinite'' and error-free data. Consequently,
deriving meaningful insights about the estimand from finite data becomes
impossible \citep[pp.~5]{Schuessler_et_al_2023}. Therefore, to ensure
accurate and reliable estimates, researchers prioritize estimators with
desirable identification properties. For instance, the Z-test is a
widely used estimator for comparing group proportions, yielding accurate
estimates when its underlying assumptions are satisfied
\citep{Kanji_2006}. Furthermore, researchers can interpret estimates
from the Z-test as causal, provided the data is collected through a
randomized experiment.

@@

Even though counterfactuals are unobservable, researchers can still
identify the ACE from associational estimates by leveraging the
structural approach. The approach identifies the ACE by statistically
conditioning data on a \emph{sufficient adjustment set} of variables
\(X\) \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. This
\emph{sufficient} set (potentially empty) must block all non-causal
paths between \(T\) to \(Y\) without opening new ones. When such a set
exists, then \(T\) and \(Y\) are \emph{d-separated} by \(X\)
(\(T \:\bot\:Y \:|\:X\)) \citep{Pearl_2009}, and \(X\) satisfies the
\emph{backdoor criterion} \citep[pp 37]{Neal_2020}. Here,
\emph{conditioning} describes the process of restricting the focus to
the subset of the population defined by the conditioning variable
\citep[pp.~32]{Neal_2020} (see Equation~\ref{eq-CACE}).

@@

For instance, if researchers cannot conduct the randomized experiments
described in Section~\ref{sec-appendix-A} and must instead rely on
observational data, they can still identify the ACE as long as an
observed variable \(X\), such as the socio-economic status of the
school, satisfies the backdoor criterion. Under these circumstances,
researchers first identify the \emph{conditional average causal effect}
(CACE, Equation~\ref{eq-CACE})

@@

SCMs and DAGs support causal inference through two key advantages.
First, regardless of complexity, they can represent various causal
structures using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
decompose complex structures into manageable components, facilitating
their analysis \citep{McElreath_2020}. Second, they depict causal
relationships in a non-parametric and fully interactive way. This
flexibility enables feasible ACE identification strategies without
defining the variables' data types, the functional form between them, or
their parameters \citep[pp.~35]{Pearl_et_al_2016}.
Section~\ref{sec-appendix-B21} and Section~\ref{sec-appendix-B22}
elaborate on the first advantage, while Section~\ref{sec-appendix-B22}
and Section~\ref{sec-appendix-B23} do so for the second.

@@

The structural approach to causal inference translates probabilistic
insights into actionable strategies seeking to identify the ACE from
associational quantities. The approach achieves this by relying on the
\emph{modularity assumption}, which posits that intervening on a node
alters only the causal mechanism of that node, leaving others unchanged
\citep[pp.~34]{Neal_2020}.

@@

With these new insights, researchers revisiting the scenario in
Section~\ref{sec-appendix-B22} can infer that the socio-economic status
of the school, \(X\), satisfies the backdoor criterion, assuming the
causal structure depicted by the SCM \ref{fig-scm_example2} and DAG
\ref{fig-mdag_example2} is correct. This means that \(X\) serves as a
sufficient adjustment set, as it effectively blocks all confounding
non-causal paths introduced by the fork structure. Nevertheless, since
\(Y\) remains dependent on \(T\) even after conditioning
\((Y \:\not\bot\:T \:|\:X)\), this dependency can only be attributed to
the direct causal effect \(T \rightarrow Y\). Notably, for the purpose
of identification, the conditioned DAG \ref{fig-mdag_example2} is
equivalent to the manipulated DAG \ref{fig-mdag_example3}, because \(X\)
satisfies the backdoor criterion.

@@

Researchers can then apply the \emph{backdoor adjustment} to identify
the ACE of \(T\) on \(Y\). They achieve this by first identifying the
CACE of \(T\) on \(Y\) by conditioning on \(X\), and then marginalizing
this effect over \(X\) to obtain the ACE. This process is expressed in
Equation~\ref{eq-mACE2} (see Section~\ref{sec-appendix-B}).

@@

Nevertheless, researchers can manage the complexity of the likelihood by
assuming specific local (in)dependencies among variables. SCMs and DAGs
provide a formal framework to represent these assumptions, as detailed
in Section~\ref{sec-appendix-B22}. These assumptions improve model
tractability and simplify the estimation process by enabling the
derivation of the BNF of the likelihood (Equation~\ref{eq-like_BNF}).
With this simplified structure, any probabilistic programming language
can model the system and compute the parameter's posterior distribution
using Equation~\ref{eq-bayes_rule}.
